# Building a DAG

```{r}
#| output: false
library(simDAG)
library(tidyverse)
library(dplyr)
library(dagitty)
library(tidyr)
library(stringr)
library(ggdag)
library(ggplot2)

source("./r/plot_dag.R")
source("./r/rtruncnorm.R")
```

## Geology example

In mineral resource estimation, the use of machine learning comes in handy when dealing with missing values or with dealing with highly under-sampled data such as the bulk density of rock samples. For bulk density a common approach is to simply use assays to predict density. Assays have molar mass which is directly correlated with bulk density ($assays \rightarrow bulk\ density$) of a rock sample.

To simplify the visual presentation of the DAG, the DAG nodes will be abbreviated.

DAG node abbreviations:

- X = assays
- Y = rock bulk density
- L = lithology
- M = mineral assemblage
- R1 = porosity
- R2 = moisture
- R3 = texture
- R4 = hardness
- R5 = rock competence
- G1 = weathering
- G2 = alteration
- G3 = mineralisation
- G4 = oxidation

```{r}
#| label: fig-dag-geo1
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 6
#| fig.height: 4
dagify(
  R1 ~ G1 + G2 + G3 + R3,
  R2 ~ G1,
  R3 ~ L + G2,
  R4 ~ L + G1 + G2,
  R5 ~ G1 + R1 + R3,
  G4 ~ G1,
  M ~ L + G1 + G2 + G3 + G4,
  X ~ M,
  Y ~ X + M + R1 + R2
) |> plot_dag(layout="stress")
```

```{r}
#| label: fig-dag-geo2
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 6
#| fig.height: 4
dagify(
  M ~ L + G,
  R ~ L + G,
  X ~ M,
  Y ~ X + M + R,
  S ~ X + G + L,
  labels = c(
    M = "Mineral Assemblage",
    L = "Lithology",
    G = "Geological Processes",
    R = "Rock Properties",
    X = "Assays",
    Y = "Bulk Density",
    S = "Domain"
  )
) |> plot_dag(layout="stress") +
  geom_dag_label_repel(
    aes(label=label),
    segment.alpha = 1,
    size = 3,
    show.legend = FALSE,
    box.padding = grid::unit(2, "lines"),
    label.padding = grid::unit(0.2, "lines")
  )
```

## Geometallurgy example

```{r}
offsets_x <- c("1"=1.52,"2"=0.64,"3"=10.66,"4"=0.64,
               "5"=0.82,"6"=22.9,"7"=1.13)
sigma_x <- c("1"=1.15,"2"=0.39,"3"=4.52,"4"=0.01,
             "5"=0.63,"6"=12.57,"7"=0.81)
node_x <- function(data, parents, betas=NULL) {
  m <- offsets_x[data$U]
  z <- rtruncnorm(nrow(data), mean=m, sd=sigma_x[data$U], a=0.001)
  return(z)
}

offsets_y <- c("1"=82.65,"2"=63.31,"3"=82.55,"4"=35.08,
               "5"=48.64,"6"=91.58,"7"=73.04)
slope_y <- c("1"=0.78,"2"=0.16269347,"3"=0.02,"4"=0.01,
             "5"=4.44,"6"=-0.14,"7"=0.42)
sigma_y <- c("1"=3.24,"2"=2.6,"3"=1.99,"4"=0.01,
             "5"=3.24,"6"=1.9,"7"=2.88)
node_y <- function(data, parents, betas=NULL) {
  m <- offsets_y[data$U] + slope_y[data$U]*data$X
  z <- rtruncnorm(nrow(data), mean=m, sd=sigma_y[data$U], a=0.001)
  return(z)
}

dag_geomet <- empty_dag() +
  node("U", type="rcategorical", probs=c(0.64,0.07,0.02,0.01,0.02,0.01,0.23),
       output="factor", labels=1:7) +
  node("Z", type="rnorm", mean=0, sd=1) +
  node("X", type=node_x, parents="U") +
  node("Y", type=node_y, parents="U")

# Simulate the Naive Measured Dataset
set.seed(1700)
df_geomet <- sim_from_dag(dag_geomet, n_sim=100) |> 
  as.data.frame() |>
  mutate(Source = "Measured") |>
  select(X, Y)
```

The synthesized recovery on grade is shown in @fig-geomet1.

```{r}
#| label: fig-geomet1
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 5
#| fig.height: 5

v <- densCols(df_geomet, colramp=colorRampPalette(c("black", "white")))
v <- col2rgb(v)[1,] + 1L
cols <- viridis::viridis(256, alpha=1)
cols <- cols[v]

plot(df_geomet$X, df_geomet$Y, ylim=c(0,100), col=cols, pch=16, xlim=c(0.01,15),
     main="Synthesized Geomet data", xlab="Grade (g/t)", ylab="Metal Recovery (%)")

xx <- c(0.001, seq(0.01, 25, by=0.01))
yy <- 91*(1-exp(-20*xx)) + 0.9*log(xx)
lines(xx, yy, lty="dashed", col="black")
yy <- 84*(1-exp(-5*xx)) + 1*log(xx)
lines(xx, yy, lty="solid", col="black", lwd=2)
yy <- 75*(1-exp(-0.6*xx)) + 2*log(xx)
lines(xx, yy, lty="dashed", col="black")
```

## Crafting a DAG from a list of mineral formula

```{r}
classify_mineral_group <- function(formula) {
  case_when(
    # Native Minerals: single element symbol (e.g., Au, Cu, C, Ag)
    str_detect(formula, "^[A-Z][a-z]?[0-9]*$") ~ "Native Minerals",
    
    # Silicates: contains Silicon (Si)
    str_detect(formula, "Si") ~ "Silicates",
    
    # Carbonates: contains CO3
    str_detect(formula, "CO3") ~ "Carbonates",
    
    # Phosphates: contains PO4 or P with O
    str_detect(formula, "PO4|P") ~ "Phosphates",
    
    # Sulphates: contains SO4 group
    str_detect(formula, "SO4") ~ "Sulphates",
    
    # Sulphides: contains S (without SO4 group)
    str_detect(formula, "S") ~ "Sulphides",
    
    # Halides: contains Halogens (F, Cl, Br, I) without Oxygen
    str_detect(formula, "F|Cl|Br|I") & !str_detect(formula, "O") ~ "Halides",
    
    # Oxides & Hydroxides: contains Oxygen (O/OH) not captured above
    str_detect(formula, "O|OH") ~ "Oxides",
    
    TRUE ~ "Other"
  )
}

create_mineral_dag <- function(data) {
  pattern <- "H2O|CO3|CO2|OH|[A-Z][a-z]?"
  
  # Clean names & classify into Mineral Groups
  processed_df <- data %>%
    mutate(
      mineral_clean = str_replace_all(mineral, "[^a-zA-Z0-9_]", "_"),
      mineral_group = classify_mineral_group(formula)
    )
  
  # Layer 1 Edges (Mineral Group -> Mineral)
  group_edges <- processed_df %>%
    distinct(mineral_group, mineral_clean) %>%
    mutate(
      group_clean = str_replace_all(mineral_group, "[^a-zA-Z0-9_]", "_"),
      edge_str = paste0("  ", group_clean, " -> ", mineral_clean)
    )
  
  # Layer 2 Edges (Mineral -> Element / Polyatomic Node)
  element_edges <- processed_df %>%
    mutate(nodes = str_extract_all(formula, pattern)) %>%
    unnest(cols = c(nodes)) %>%
    distinct(mineral_clean, nodes) %>%
    mutate(edge_str = paste0("  ", mineral_clean, " -> ", nodes))
  
  # Combine Edges into DOT Notation
  all_edges <- c(group_edges$edge_str, element_edges$edge_str)
  dot_notation <- paste0("dag {\n", paste(all_edges, collapse = "\n"), "\n}")
  
  # Parse into dagitty object
  dag_obj <- dagitty(dot_notation)
  
  return(list(
    dot = dot_notation,
    data = processed_df,
    dag = dag_obj
  ))
}
```

Now for an example.

```{r}
# Test dataset across all 8 mineral groups
df <- data.frame(
  mineral = c("Quartz", "Hematite", "Calcite", "Galena", 
              "Gypsum", "Halite", "Apatite", "Gold", "Malachite"),
  formula = c("SiO2", "Fe2O3", "CaCO3", "PbS", 
              "CaSO4·2H2O", "NaCl", "Ca5(PO4)3F", "Au", "Cu2(OH)2CO3")
)

result <- create_mineral_dag(df)

# Print generated DOT string
cat("=== 3-Tier DOT Notation Output ===\n")
cat(result$dot)
cat("\n==================================\n\n")
```

The resulting DAG is shown in @fig-min-dag.

```{r}
#| label: fig-min-dag
#| fig-cap: "Example DAG from mineral formula"
#| fig.width: 8
#| fig.height: 6

# --- Plot ---
tidy_dagitty(result$dag, layout="stress") |>
  ggplot(aes(x=x, y=y, xend=xend, yend=yend)) +
  geom_dag_edges_link(aes(start_cap=ggraph::circle(2,"mm"),
                          end_cap=ggraph::circle(2,"mm"))) +
  geom_dag_label(aes(label=name), 
                       color="black", 
                       fill="lightblue",
                       size=3) +
  theme_dag() +
  labs(
    title = "3-Tier Mineral Relationship DAG",
    subtitle = "Mineral Group  →  Mineral Species  →  Chemical Node"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray30", size = 10)
  )
```
