# Mixture of Gaussian Bayesian Network trick

Non-linear relationships and non-Gaussian distributions can be handled in a mixture of Gaussian Bayesian Network.

```{r}
#| output: false
library(bnlearn)
library(simDAG)
```

## Uniform distribution

```{r}
#| label: fig-ud1
#| fig-cap: "Approximating a Uniform distribution (bnlearn)"
#| fig.width: 4
#| fig.height: 4

set.seed(1781)

# Generate data
k <- 6
n_samples <- 1000
x <- runif(n_samples)
s <- cut(x, breaks=0:k/k)
z <- data.frame(X=x, S=s)

# Fit Bayesian Network
dag <- "[S][X|S]"
dag <- model2network(dag)
bn <- bn.fit(dag, z, replace.unidentifiable=TRUE)

# --- Plot ---
zz <- rbn(bn, n_samples)$X
plot(ecdf(z$X), lwd=2, main="ecdf", do.points=FALSE)
plot(ecdf(zz), col="red", lwd=2, do.points=FALSE, add=TRUE)
```

To replicate the Bayesian Network using simDAG we need to add the independent Gaussian random variable $Z$. This variable is a parent to $X$. Making the DAG $S \rightarrow X \leftarrow Z$.

```{r}
offsets <- as.numeric(coef(bn$X)) |>
  (\(x) {names(x) <- 1:length(x); x})()
sigma <- as.numeric(sigma(bn$X)) |>
  (\(x) {names(x) <- 1:length(x); x})()
node_x <- function(data, parents, betas=NULL) {
  return(offsets[data$S] + sigma[data$S]*data$Z)
}

dag_replicated <- empty_dag() +
  node("S", type="rcategorical", probs=coef(bn$S),
       output="factor", labels=1:length(coef(bn$S))) +
  node("Z", type="rnorm", mean=0, sd=1) +
  node("X", type=node_x, parents=c("S","Z"))

# Simulate the Naive Measured Dataset
set.seed(1781)
n_samples <- 1000
df_replicated <- sim_from_dag(dag_replicated, n_sim=n_samples) |> 
  as.data.frame()
```

```{r}
#| label: fig-ud2
#| fig-cap: "Approximating a Uniform distribution (simDAG)"
#| fig.width: 4
#| fig.height: 4

set.seed(1781)

# Generate data
k <- 6
n_samples <- 1000
x <- runif(n_samples)
s <- cut(x, breaks=0:k/k)
z <- data.frame(X=x, S=s)

# --- Plot ---
plot(ecdf(df_replicated$X), lwd=2, main="ecdf", do.points=FALSE)
plot(ecdf(zz), col="red", lwd=2, do.points=FALSE, add=TRUE)
```

## Sigmoid function

```{r}
#| label: fig-sf1
#| fig-cap: "Approximating a Sigmoid Function (bnlearn)"
#| fig.width: 4
#| fig.height: 4

sigmoid <- function(x, alpha=1) {
  y <- 1 / (1 + exp(-alpha * x))
  return(y)
}

# Generate data
n_samples <- 1000
set.seed(1781)
x <- seq(from=-5, to=5, by=0.1)
y <- sigmoid(x)
z <- data.frame(X=c(-10, x, 10), Y=c(0, y, 1))

dag <- "[S][R|S][X|S:R][Y|X:S]"
dag <- model2network(dag)

# Fit Bayesian Network
zz <- z
zz$S <- cut(zz$X, breaks=10)
zz$R <- cut(zz$Y, breaks=10)
bn <- bn.fit(dag, zz, replace.unidentifiable=TRUE)

# --- Plot ---
x <- z
x$Y <- as.numeric(NA)
x$S <- cut(x$X, breaks=nlevels(zz$S))
x$R <- factor(NA, levels=levels(zz$R))
y <- suppressWarnings(impute(bn, x))

plot(z$X, z$Y, lwd=2, type="l", main="Sigmoid Function",
     xlab="X", ylab="Y = sigmoid(X)")
lines(y$X, y$Y, type="l", col="red", lwd=2)
```

## Reversing an edge

```{r}
#| label: fig-re1
#| fig-cap: "Reversing an edge"
#| fig.width: 4
#| fig.height: 4

set.seed(1781)

# Generate data
k <- 11
n_samples <- 2000
A <- rnorm(n_samples)
B <- as.factor(ifelse(A>0.0, "b", "nb"))
s <- cut(A, breaks=k)
z <- data.frame(A=A, B=B, S=s)

# Fit Bayesian Network
dag <- "[B][S|B][A|B:S]"
dag <- model2network(dag)
bn <- bn.fit(dag, z, replace.unidentifiable=TRUE)

# --- Plot ---
y <- suppressWarnings(bnlearn::rbn(bn, n_samples))
qqnorm(y$A, main="Q-Q Plot", pch=19, cex=0.6, ylab="Sample Quantiles (A)")
```
