# Simulating data from a DAG

Simulating synthetic data directly from a Directed Acyclic Graph (DAG) allows us to create controlled testing environments. By explicitly defining structural equations, we can model how variables interact.

```{r}
#| output: false
library(simDAG)
library(bnlearn)
library(SEMgraph)
library(ggplot2)
library(ggdag)
library(patchwork)
library(dplyr)
library(tidyr)
library(broom)
source("./r/rtruncnorm.R")
source("./r/plot_dag.R")
```

## Truth dataset

The truth DAG (@fig-dag) is used to define the structure while the parameters of the probabilistic graphical model are set arbitrarily.

```{r}
dag_truth <- empty_dag() +
  node("C", type="rcategorical", probs=c(0.15, 0.47, 0.38),
       output="factor", labels=c("a", "b", "c")) +
  node("I", type="rnorm", mean=15.0, sd=1.5) +
  node("J", type="rbernoulli", p=0.5, output="logical") +
  node("L", type="rnorm", mean=-1.3, sd=5.5) +
  node("E", type="gaussian",
       formula= ~ 2.0 + I*1.5 + Cb*-1.1 + Cc*7,
       error=0.7,
       parents=c("I","C")) +
  node("G", type="gaussian", error=4.5,
       formula=~ 3.5 + Cb*5.2 + Cc*-11.1,
       parents="C") +
  node("S", type="identity",
       formula=~ (E + L) < 20.7,
       parents=c("E","L")) +
  node("M", type="gaussian",
       formula=~ 1.7 + E*3.44,
       error=3.7,
       parents="E") +
  node("D", type="gaussian",
       formula= ~ 0.1 + E*15.2 + M*-2.7 + L*4.3 + G*8.6 + JTRUE*-5.68,
       error=3.1,
       parents=c("E","M","L","G","J")) +
  node("Z", type="identity",
       formula=~ ifelse(D > 1.2, TRUE, FALSE),
       parents="D")

# Simulate the True Dataset
set.seed(2131)
n_samples <- 10000
df_truth <- sim_from_dag(dag_truth, n_sim=n_samples) |> 
  as.data.frame() |>
  mutate(Source = "Truth")
```

The results of simulating data from @fig-dag is shown in @fig-dag1. The plot shows the data generating DAG on the left hand side and a scatter plot of $D$ on $E$ where data points are coloured by the confounder variable $C$.

```{r}
#| label: fig-dag1
#| fig-cap: "Example simulating from a DAG using simDAG"
#| fig.width: 8
#| fig.height: 4

x <- df_truth[order(df_truth$S), ]

# --- Plots ---
# Left: The DAG
dag_a <- dagify(
  E ~ I + C,
  G ~ C,
  S ~ E + L,
  M ~ E,
  D ~ E + M + L + G + J,
  Z ~ D
) |> 
  tidy_dagitty(layout="stress") |>
  ggplot(aes(x=x, y=y, xend=xend, yend=yend)) +
    geom_dag_edges_link(aes(start_cap=ggraph::circle(5,"mm"),
                            end_cap=ggraph::circle(5,"mm"))) +
    geom_dag_point(size=9) +
    geom_dag_text(size=4) +
    theme_dag() +
    labs(title="DAG")

# Right: Scatter plot
plot_a <- ggplot(x, aes(x=E, y=D)) +
  geom_point(aes(color=C), alpha=0.5, size=0.5) +
  geom_smooth(method="lm", se=FALSE, color="black", linetype="dashed", size=1.2) +
  geom_smooth(aes(color=C), method="lm", se=FALSE, size=1) +
  theme_minimal() +
  labs(title="Scatter Plot: D ~ E",
       subtitle="Dashed line = Aggregate\nColored lines = Stratified by C")

# Display side-by-side
dag_a + plot_a + plot_layout(widths=c(8, 6))
```

### The measured dataset

The truth dataset (`df_truth`) is subset by filtering rows where the variable $S$ is true. Only the variables $E$, $D$, and $Z$ are selected. The $Z$ variable is converted from logical to a factor.

```{r}
df_measured <- df_truth[df_truth$S, c("E","D","Z")] |>
  mutate(Z = as.factor(Z),
         Source = "Truth")
```

### The replicated dataset

```{r}
# Learn the DAG and fit the Bayesian Network
dag <- hc(df_measured[,-4])
bn_model <- bn.fit(dag, df_measured[,-4])

# Generate synthetic data
set.seed(101)
n_samples <- 2400
df_replicated <- rbn(bn_model, n=n_samples) |> 
  mutate(Source = "Synthetic")
```

Comparing the replicated dataset generated from the Bayesian Network against the true measured dataset shows very good agreement between the two datasets (@fig-res1).

```{r}
#| label: fig-res1
#| fig-cap: "Comparison between measured data and synthetic data"
#| fig.width: 6
#| fig.height: 3

# Select common continuous variables for comparison
continuous_vars <- c("D", "E", "Z")

combined_df <- bind_rows(
  df_measured |> select(all_of(continuous_vars), Source) |> mutate(Z = as.numeric(Z)),
  df_replicated |> select(all_of(continuous_vars), Source) |> mutate(Z = as.numeric(Z))
) |>
  pivot_longer(cols=all_of(continuous_vars), names_to="Variable", values_to="Value")

# Plot using ggplot2
ggplot(combined_df, aes(x=Value, color=Source)) +
  stat_ecdf(geom="step", linewidth=1) +
  facet_wrap(~ Variable, scales="free_x") +
  scale_color_manual(values=c("darkorange", "royalblue")) +
  labs(
    title = "Empirical Cumulative Distribution Function (ECDF)",
    subtitle = "Evaluating structural shifts introduced by missing features",
    x="Measured Value",
    y="Cumulative Probability (Fn(x))",
    color="Dataset Version"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text=element_text(face="bold", size=11),
    plot.title=element_text(face="bold", size=14)
  )
```

```{r}
#| label: fig-res2
#| fig-cap: "Comparison between truth data and synthetic data"
#| fig.width: 6
#| fig.height: 3

# Select common continuous variables for comparison
continuous_vars <- c("D", "E", "Z")

combined_df <- bind_rows(
  df_truth |> select(all_of(continuous_vars), Source) |> mutate(Z = as.numeric(Z)),
  df_replicated |> select(all_of(continuous_vars), Source) |> mutate(Z = as.numeric(Z))
) |>
  pivot_longer(cols=all_of(continuous_vars), names_to="Variable", values_to="Value")

# Plot using ggplot2
ggplot(combined_df, aes(x=Value, color=Source)) +
  stat_ecdf(geom="step", linewidth=1) +
  facet_wrap(~ Variable, scales="free_x") +
  scale_color_manual(values=c("darkorange", "royalblue")) +
  labs(
    title = "Empirical Cumulative Distribution Function (ECDF)",
    subtitle = "Evaluating structural shifts introduced by missing features",
    x="True Value",
    y="Cumulative Probability (Fn(x))",
    color="Dataset Version"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text=element_text(face="bold", size=11),
    plot.title=element_text(face="bold", size=14)
  )
```

```{r}
#| label: fig-compare-dags
#| fig-cap: "Truth DAG"
#| fig.width: 6
#| fig.height: 4

dag_true <- dagify(
  E ~ I + C,
  G ~ C,
  S ~ E + L,
  M ~ E,
  D ~ E + M + L + G + J,
  Z ~ D
) |> plot_dag(layout="stress") +
  labs(title="True DAG")

dag_bn <- as.igraph(bn_model) |>
  graph2dagitty() |>
  plot_dag(layout="stress") +
  labs(title="Bayesian Network DAG")

# Display side-by-side
dag_true + dag_bn
```
