# Simpson's paradox (The Confounding/Mediation Pivot)

Simpson's paradox occurs when a statistical trend appears in different groups of data but disappears or reverses when these groups are combined.

## Simulating data from a DAG

```{r}
#| output: false
library(simDAG)
library(bnlearn)
library(SEMgraph)
library(ggplot2)
library(ggdag)
library(patchwork)
library(dplyr)
library(tidyr)
library(broom)
```

```{r}
#| label: fig-sp1
#| fig-cap: "Simpson's Paradox - classic example"
#| fig.width: 8
#| fig.height: 4

set.seed(123)

# Generate data
n <- 300
z <- sample(1:3, n, replace=TRUE)
x <- 2 * z + rnorm(n, mean=0, sd=0.5)
y <- 5 * z - 1.5 * x + rnorm(n, mean=0, sd=0.5)
simpson_data <- data.frame(X=x, Y=y, Z=as.factor(z))

# --- Plots ---
# Left: The DAG
dag_simpson <- dagify(X ~ Z, Y ~ Z, Y ~ X, exposure="X", outcome="Y") |>
  tidy_dagitty() |>
  ggdag(node_size=12) + 
  theme_dag() + 
  labs(title = "Simpson's Paradox DAG")

# Right: The Scatter Plot
plot_simpson <- ggplot(simpson_data, aes(x=X, y=Y)) +
  geom_point(aes(color=Z), alpha=0.7) +
  geom_smooth(method="lm", se=FALSE, color="black", linetype="dashed", size=1.2) +
  geom_smooth(aes(color=Z), method="lm", se=FALSE, size=1) +
  theme_minimal() +
  labs(title="Scatter Plot: Y ~ X",
       subtitle="Dashed line = Aggregate (Omitted Z)\nColored lines = Stratified by Z")

# Display side-by-side
dag_simpson + plot_simpson
```

```{r}
#| label: fig-sp2
#| fig-cap: "Simpson's Paradox - more difficult example"
#| fig.width: 8
#| fig.height: 4

set.seed(123)

n_group <- 150

# Group 1: High Z, lower X, steep negative slope
z1 <- data.frame(
  X = rnorm(n_group, mean=2, sd=0.8),
  Z = "Group 1"
) |> mutate(Y = 3 + 4.5 * X + rnorm(n_group, sd=4.8))

# Group 2: Medium Z, medium X, flat/slight positive slope
z2 <- data.frame(
  X = rnorm(n_group, mean=4, sd=0.8),
  Z = "Group 2"
) |> mutate(Y = 5 + 0.2 * X + rnorm(n_group, sd=4.8))

# Group 3: Low Z, higher X, moderate positive slope
z3 <- data.frame(
  X=rnorm(n_group, mean=6, sd=0.8),
  Z="Group 3"
) |> mutate(Y = 0.2 + 0.8 * X + rnorm(n_group, sd=0.8))

# Combine into a single dataset
simpson_subtle <- rbind(z1, z2, z3)

# --- Plots ---
# Left: The DAG
dag_simpson <- dagify(X ~ Z, Y ~ Z, Y ~ X, exposure="X", outcome="Y") |>
  tidy_dagitty() |>
  ggdag(node_size=12) + 
  theme_dag() + 
  labs(title="Simpson's Paradox DAG")

# Right: The Scatter plot
plot_simpson <- ggplot(simpson_subtle, aes(x=X, y=Y)) +
  geom_point(aes(color=Z), alpha=0.4) +
  scale_color_brewer(palette="Set1") +
  geom_smooth(method="lm", se=FALSE, color="black", linetype="dashed", size=1.2) +
  geom_smooth(aes(color=Z), method="lm", se=FALSE, size=1) +
  theme_minimal() +
  labs(
    title="Subtle Simpson's Paradox (Overlapping Groups)",
    subtitle="Dashed line = Combined Positive Trend\nColored lines = True Group Slopes",
    color="Confounder (Z)"
  )

# Display side-by-side
dag_simpson + plot_simpson
```

## Automatically detecting

The function `detect_simpsons_paradox` fits a global linear model ($y∼x$) and records the slope sign (+ or −). It then splits the data by your designated confounder and recalculates the slopes. If any subgroup slope sign opposes the global slope sign, it flags the Simpson's paradox.

```{r}
# Detect Simpson's Paradox
# Compares global regression slope vs. stratified subgroup regression slopes
detect_simpsons_paradox <- function(data, x_var, y_var, group_var) {
  message("\n--- Checking for Simpson's Paradox ---")
  
  # Global Model
  global_formula <- as.formula(paste(y_var, "~", x_var))
  global_slope <- lm(global_formula, data=data) |> 
    tidy() |> 
    filter(term == x_var) |> 
    pull(estimate)
  
  # Grouped Model
  group_slopes <- data |>
    nest_by(!!sym(group_var)) |>
    mutate(model = list(lm(global_formula, data = data))) |>
    summarise(
      Group = !!sym(group_var),
      Slope = tidy(model) |> filter(term == x_var) |> pull(estimate)
    )
  
  # Print results
  print(paste("Global Linear Slope:", round(global_slope, 4)))
  print("Slopes by Subgroup:")
  print(group_slopes)
  
  # Paradox Check: Do the subgroup slopes have a different sign than the global slope?
  paradox_detected <- any(sign(group_slopes$Slope) != sign(global_slope))
  
  if (paradox_detected) {
    message("⚠️ SIMPSON'S PARADOX DETECTED! The global trend contradicts the subgroup trends.")
  } else {
    message("✅ No Simpson's Paradox detected. Trends are consistent.")
  }
}
```

Example

```{r}
set.seed(42)
n_samples <- 1000

# Lithology A: Weathered/Porous Ore (High Iron, Low Bulk Density due to vugs/voids)
lith_A <- tibble(
  Lithology    = "Weathered_Ore",
  Iron_Pct     = rnorm(n_samples / 2, mean = 55, sd = 5),
  # Within this lithology, higher Fe means slightly higher density
  Bulk_Density = 2.0 + (0.02 * Iron_Pct) + rnorm(n_samples / 2, sd = 0.1),
  Gold_g_t     = rnorm(n_samples / 2, mean = 0.8, sd = 0.3),
  Copper_Pct   = rnorm(n_samples / 2, mean = 0.4, sd = 0.15)
)

# Lithology B: Fresh BIF (Lower Iron overall, but dense, tight mineral matrix)
lith_B <- tibble(
  Lithology    = "Fresh_BIF",
  Iron_Pct     = rnorm(n_samples / 2, mean = 35, sd = 5),
  # Within this lithology, higher Fe also means higher density
  Bulk_Density = 3.2 + (0.02 * Iron_Pct) + rnorm(n_samples / 2, sd = 0.1),
  Gold_g_t     = rnorm(n_samples / 2, mean = 0.7, sd = 0.3),
  Copper_Pct   = rnorm(n_samples / 2, mean = 0.35, sd = 0.15)
)

# Combine
deposit_df <- bind_rows(lith_A, lith_B) |>
  mutate(
    # Engineering Collider: Stockpile Selection Rule
    # Core is sent to High-Grade stockpile if Cu > 0.6% OR Au > 1.2 g/t
    In_High_Grade_Stockpile = ifelse(Copper_Pct > 0.6 | Gold_g_t > 1.2, "Yes", "No")
  )

# Test 1: Density vs Iron, grouped by Lithology (Confounder)
detect_simpsons_paradox(
  data = deposit_df, 
  x_var = "Iron_Pct", 
  y_var = "Bulk_Density", 
  group_var = "Lithology"
)
```

```{r}
#| fig-cap: "Example Simpsons Paradox"
#| label: fig-dag
#| fig-width: 7
#| fig-height: 4

# Quick Visual Verification
# Simpson's Visual Validation
ggplot(deposit_df, aes(x = Iron_Pct, y = Bulk_Density)) +
  geom_point(aes(color = Lithology), alpha = 0.4) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") + # Global
  geom_smooth(aes(color = Lithology), method = "lm", se = FALSE) + # Grouped
  labs(title = "Simpson's Paradox Visualized", subtitle = "Dotted black line = Global, Colored lines = Per-Lithology") +
  theme_minimal()
```
